---
title: "Nonlinear Regression Agricultural Applications"
author: "Fernando Miguez"
date: "`r Sys.Date()`"
fig_width: 6
fig_height: 4
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{nlraa: An R package for Nonlinear Regression Applications in Agricultural Research}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 7, fig.height = 6)
library(ggplot2)
library(nlraa)
```

# Nonlinear Regression

For an introduction to this topic see the publication by Archontoulis and Miguez (https://doi.org/10.2134/agronj2012.0506) and the book chapter by Miguez, Archontoulis and Dokoohaki (https://doi.org/10.2134/appliedstatistics.2016.0003).

One of the objectives of those publications was to introduce a large family of nonlinear functions to practicioners in the agricultural research area to a variety of tools that can use to fit data which does not conform to a linear relationship. The feature that distinguishes this approach from others such as ploynomials, splines or gams (to name a few) is that the parameters of the model have biologically meaningful interpretations. In R the approach that makes fitting nonlinear mixed models almost as easy as fitting linear mixed models is the use of self starting functions.

# Index of self starting functions

## Functions in 'base' R

**stats** package

1. SSasymp (Asymptotic)
2. SSasympOff (Asymptotic with an offset)
3. SSasympOrig (Asymptotic through the Origin)
4. SSbiexp (Bi-exponential)
5. SSfol (First order compartment)
6. SSfpl (Four parameter logistic)
7. SSgompertz (Gompertz)
8. SSlogis (Logistic)
9. SSmicmen (Michaelis-Menten)
10. SSweibull (Weibull)

## Functions in this package (nlraa)

1. SSbgf (Beta-Growth Function)
2. SSbgf4 (Four Parameter Beta-Growth Function)
3. SSdlf (Declining Logistic Function)
4. SSricker (Ricker population growth)
5. SSprofd (Profile of protein distribution)
6. SSlinp (linear-plateau)
7. SSplin (plateau-linear)
8. SSquadp (quadratic-plateau)
9. SSpquad (plateau-quadratic)
10. SSblin (bilinear)
11. SSexpf (exponential function)

For the reference on the Beta growth function see Yin et al. 2003 (https://doi.org/10.1093/aob/mcg029) and Erratum: https://doi.org/10.1093/aob/mcg091.
For a reference on a use of the declining logistic see Oddi et al. (2019). For the Ricker model see: https://en.wikipedia.org/wiki/Ricker_model. For the protein profile in the canopy see Johnson et al. (2010) doi:10.1093/aob/mcq183. 

The main benefit of having these self starting functions is that they can be incorporated in the workflow of package 'nlme' which allows for fitting nonlinear mixed models. 

## Which functions are available?

```{r apropos}
apropos("^SS")
```

# Datasets

These are the available datasets in the package (so far):

1. "sm"
2. "lfmc"
3. "swpg"

```{r sm}
## Sorghum and Maize dataset
data(sm)
ggplot(data = sm, aes(x = DOY, y = Yield, color = Crop)) + 
  geom_point() + 
  facet_wrap(~ Input)
```

```{r lfmc}
## Live fuel moisture content
data(lfmc)
ggplot(data = lfmc, aes(x = time, y = lfmc, color = leaf.type)) +
  geom_point() + 
  ylab("Live fuel moisture content (%)")
```

```{r swpg}
## Soil water and plant growth
data(swpg)
ggplot(data = swpg, aes(x = ftsw, y = lfgr)) +
  geom_point() + 
  xlab("Fraction Transpirable Soil Water") + 
  ylab("Relative Leaf Growth")
```

# What to do when convergence fails?

Convergence problems in nonlinear models are often because of a number of reasons:

1. The model is not appropriate for the observed data (or viceversa)
2. The model is conceptually correct but there is an error in the formula
3. The model is too complex; a simpler model should be used
4. The model is too simple; a more complex model should be used
5. Starting values are too far from the solution

Model specification (choosing the correct model) is clearly very important when using 
nonlinear models. For convergence problems related to poor starting values there are some alternatives:

1. Try algorithm 'port' in function 'nls'
2. Use an alternative algorithm 'Levenberg-Marquardt' in package 'minpack.lm' through the function 'nlsLM', which can be more robust
3. Use function 'nls2' in package 'nls2' which uses a 'brute-force' approach of searching over a grid
4. Define a custom optimization and use function 'optim' in 'stats' package (also 'nlm' or 'nlminb'). This option extends the possibility of available algorithms.
5. Manually construct a profile of residual sum of squares as a function of the parameters to understand the relationship which might lead to a better choice for starting values.

## Other modeling approaches

Other modeling approaches do not use such a direct rigid specification as the models classically used in nonlinear relationships. Some examples are:

1. **splines** (packages 'splines' or 'splines2')
2. **gams** (package 'mgcv')
3. **loess** (package 'stats')
4. **quantile regression** (packages 'quantreg*')
5. **polynomials** (poly in package 'stats')
6. **segmented regression** (segmented in package 'segmented')

Although these previous methods are much more flexible than classical nonlinear regression, the traditional approaches have the benefit of being simple and providing parameters with a straight-forward interpretation.

# Contributed packages

I have not tested any of these packages. They are here for reference.

## **FlexParamCurve** package

See Oswald, Nisbet, Chiaradia and Arnold, MEE, (2012) ( https://doi.org/10.1111/j.2041-210X.2012.00231.x).

1. SSposnegRichards (Positive-Negative Richards)

## **sicegar** package

See Umut Caglar, Teufel and Wilke (https://peerj.com/articles/4251/)

It was removed from CRAN because errors were not resolved. 
Not a good sign

Package still available at https://github.com/wilkelab/sicegar

## **drc** package 

Dose-response curves package

Ritz C, Baty F, Streibig JC, Gerhard D
(2015) Dose-Response Analysis Using R. PLoS ONE
10(12): e0146021. doi:10.1371/journal.pone.0146021

## **easynls** pacakge

Might make it easier to fit and plot a few different models.

## **HydroMe** package

This package has functions for fitting water retention curves. Interestingly, it
uses the 'minpack.lm' package which has a different implementation of the 
minimization algorithm (Levenberg-Marquardt).

This package has the following selfStart functions:

1. SSfredlund
2. SSgampt
3. SSgard
4. SSgardner
5. SShorton
6. SSkosugi
7. SSomuto
8. SSphilip
9. SSvgm

And additional models, such as:

1. Tani
2. Brook
3. Campbel
4. Expo

# Additional references

1. https://www.ncbi.nlm.nih.gov/pubmed/20831877
2. https://www.ncbi.nlm.nih.gov/pubmed/28582419


